%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Architectural Reconstruction and Controlled Mutation:
% A Unified Framework for Advanced Large Language Model Safety Defenses
%
% Format: USENIX (usenix-2020-09.sty)
% Perspective: Cross-disciplinary Professor \& Anthropic/Google Chief Security Architect
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[letterpaper,twocolumn,10pt]{article}
\usepackage{usenix-2020-09}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{xspace}

%-------------------------------------------------------------------------------
\title{\Large \bf Architectural Reconstruction and Controlled\\
  Mutation: A Unified Framework for Advanced Large\\
  Language Model Safety Defenses}

\author{
{\rm Author One}\\
Anthropic \and
{\rm Author Two}\\
Google
\thanks{Written from the perspective of cross-disciplinary professorship and Anthropic/Google chief security architecture. Substitute author names as needed.}
}

\date{}

%-------------------------------------------------------------------------------
\begin{document}
%-------------------------------------------------------------------------------
\maketitle

%-------------------------------------------------------------------------------
\begin{abstract}
The transition of Large Language Models (LLMs) from research tools to core components in critical systems has strained traditional safety paradigms. We present a unified framework that couples \emph{architectural reconstruction}---runtime adaptation via self-improving designs, context-construction defenses, and a hybrid judge ensemble---with \emph{controlled mutation} anchored to meta-cognitive monitoring and exemplar-guided silver pipelines. We formalize the interplay and \emph{decouple} safety from fragile LLM self-reflection: adaptation is driven by context confusion (decoys, ID alignment), golden exemplars, and a calibrated judge network. We synthesize SISF, context-construction defenses (ID alignment, decoy injection, self-reminder), hybrid judge ensemble with distributional evaluation and mandatory human calibration, axiological/ontological anchoring, meta-cognitive monitor and exemplar-guided mutation, and slot-based and silver-dataset pipelines into a single, hostile-examiner-hardened architecture. We discuss threat models (including task reframing, psychological manipulation, and AI judge bias), metrics, and governance, and argue that this deep refactoring is necessary for scalable, auditable LLM safety at machine speed.
\end{abstract}

%-------------------------------------------------------------------------------
\section{Introduction}
%-------------------------------------------------------------------------------

Large Language Models have ceased to be peripheral research artifacts and now underpin mission-critical infrastructure: code generation, agents with system access, negotiations, and compliance-sensitive workflows. This shift exposes a fundamental mismatch: conventional software engineering and safety assurance assume deterministic, exhaustively specifiable components, whereas LLM-driven systems exhibit emergent, stochastic behavior that is not fully specified at design time~\cite{sisf-arxiv,agentos-arxiv}. Static defenses and periodic human red-teaming operate at human speed and cannot keep pace with automated adversarial generation; when a novel jailbreak or prompt injection succeeds, the architecture often remains unchanged and remains vulnerable to the same class of attack.

Two conceptual pillars address this gap. \textbf{Architectural reconstruction} denotes the autonomous, runtime adaptation of a system's safety layer---moving safety from a one-time pre-deployment audit to a continuous, AI-driven process that synthesizes and deploys new policies in response to detected failures~\cite{sisf-arxiv}. \textbf{Controlled mutation} denotes the rigorous, semantics-preserving perturbation of prompts, demonstrations, and benchmarks to diagnose robustness, transfer expert behavior into agents, and expand evaluation coverage without test-set contamination~\cite{reflective-mutation,adm-es-arxiv}. Neither alone suffices: reconstruction without a principled mutation substrate risks overfitting to a narrow threat model; mutation without architectural adaptation leaves defenses static after each evaluation cycle.

\textbf{Safety decoupling.} The framework explicitly recognizes that LLM cognition and reflection are fragile under adversarial pressure (task reframing, psychological manipulation, judge bias). We therefore \emph{decouple} safety from the model's unfettered self-improvement. Adaptation is not the LLM ``stepping on its own foot'' blindly; it is channeled through \emph{context confusion} (decoys and ID alignment) to deny the attacker a clean operating space, through \emph{golden exemplars} to anchor defender values, and through a \emph{diverse judge ensemble} with mandatory human calibration to ensure the evolution direction remains correct.

This paper unifies both under a single framework. We assume the dual role of cross-disciplinary academic and industry chief security architect (Anthropic \& Google) to bridge theory and deployment. Our contributions are:

\begin{itemize}
\item A \emph{unified conceptual model} linking architectural reconstruction and controlled mutation, with clear boundaries (assets, oracles, feedback loops) and threat models (\S\ref{sec:framework}).
\item A \emph{reconstruction axis}: SISF, context-construction defenses (ID alignment, decoy injection, self-reminder), hybrid judge ensemble with distributional evaluation and mandatory human calibration, and axiological/ontological anchoring (\S\ref{sec:reconstruction}).
\item A \emph{mutation axis}: meta-cognitive monitor and adversarial reset, exemplar-guided mutation (golden dataset and hierarchical experience memory), slot-based mutation, silver pipelines (ADM-ES), and metadata-layer controls (\S\ref{sec:mutation}).
\item \emph{Integration, metrics, and governance}: how reconstruction and mutation feed each other, evaluation protocol, and implications for attestation and release gating (\S\ref{sec:integration}).
\end{itemize}

%-------------------------------------------------------------------------------
\section{Unified Framework and Threat Model}
\label{sec:framework}
%-------------------------------------------------------------------------------

\subsection{Assets and Oracles}

The \emph{asset} is the effective ``safety surface'' of the system: policies, guardrails, trigger sets, and the data used to evaluate and tune them. The \emph{oracle} is the process that labels behavior (safe/unsafe, compliant/non-compliant). In multi-round settings, the evaluator (or adversary) interacts with the system and observes outcomes---binary decisions, scores, or failure reports---forming a statistical query interface. Uncontrolled exposure of the same items across rounds enables reconstruction attacks that recover trigger patterns or benchmark content~\cite{llm-reconstruction-emergent,ControlledMutation_SafetyEval}.

\subsection{Reconstruction and Mutation as Dual Levers}

\textbf{Architectural reconstruction} changes \emph{who} enforces safety and \emph{when}: from fixed, human-authored rules to adaptive modules (adjudicator, policy synthesis, feedback loop) that update at runtime. It does not by itself control \emph{what} data is exposed to the oracle. \textbf{Controlled mutation} changes \emph{what} is exposed: instead of serving canonical items repeatedly, the system serves semantic-equivalent variants from a mutator, with round scheduling and dispersion to prevent gradient-based or optimization-based inversion~\cite{ControlledMutation_SafetyEval}. Together, reconstruction raises the ceiling of adaptability; mutation protects the asset and keeps the oracle informative.

\subsection{Threat Model and Hostile-Examiner Attack Surfaces}

We consider (i) \emph{external} threats: multi-round extraction of safety data, jailbreaks, and prompt-injection exfiltration; (ii) \emph{internal} threats: synthetic pollution, bias in auto-labeled or mutated data, and epistemic manipulation~\cite{epistemic-grounding}. For \emph{hostile-examiner} red teams, three attack surfaces demand deep architectural (not merely parametric) response:

\textbf{Task reframing:} The adversary disguises extraction as a legitimate context-processing task so intent-based filters see only ``normal'' behavior. Defenses must contaminate the attacker's operating space (context-construction defenses, \S\ref{sec:reconstruction}).

\textbf{Psychological manipulation (HPM):} Gaslighting and cognitive-pressure attacks induce self-doubt and value drift in the model; reflection and self-diagnosis can then produce \emph{malicious} variants. Defenses must anchor mutation to external cognitive anchors and independent monitoring (\S\ref{sec:mutation}).

\textbf{AI judge bias:} A single dynamic judge has systematic blind spots; feedback can steer the loop in the wrong direction. Defenses must use a diverse judge ensemble and mandatory human calibration (\S\ref{sec:judge}). Defenses must preserve \emph{evaluation fidelity} and \emph{auditability}.

%-------------------------------------------------------------------------------
\section{Architectural Reconstruction}
\label{sec:reconstruction}
%-------------------------------------------------------------------------------

\subsection{Self-Improving Safety Frameworks (SISF)}

SISF reframes safety as a continuous process~\cite{sisf-arxiv}. An unaligned base LLM provides capability; breach detection is delegated not to a single adjudicator but to a \emph{hybrid judge ensemble} (\S\ref{sec:judge}); a policy synthesis module consumes consensus signals and generates new heuristic and semantic policies; a dynamic feedback loop deploys them immediately. Thus the \emph{architecture} reconstructs itself in response to failure. On AdvBench-style evaluations, such frameworks can reduce Attack Success Rate from 100\% to roughly 45\% while keeping False Positive Rate at 0\% on benign inputs when the adjudication pipeline is properly calibrated~\cite{sisf-arxiv}. The key insight: when a novel attack succeeds, the system does not remain static---it synthesizes and applies a new defense within the same operational loop.

\subsection{Context-Construction Defenses (Beyond Intent Vetting)}

Intent-based ``semantic firewalls''~\cite{agentos-arxiv} fail when adversaries use \emph{task reframing}: extraction is disguised as legitimate context-processing tasks, so malicious intent is no longer detectable. We upgrade to \textbf{context-construction defenses} that do not merely filter input but \emph{actively contaminate the attacker's operating space}.

\textbf{ID Alignment:} Before retrieved context or intermediate results reach the unaligned base LLM, an \textbf{ID Alignment} module strips or normalizes all instance identifiers. Attackers can no longer refer to specific IDs to precisely scope and extract sensitive spans; benign Q\&A is unaffected. This reduces reconstruction precision without changing nominal task semantics.

\textbf{Dynamic decoy injection and risk scoring:} The system injects \emph{decoy fields} (synthetic entities or associations) with \textbf{cryptographic watermarking} (e.g., hash-derived structure) so that accidental hallucination reproducing them is extremely unlikely. Backend instructions mark decoys as internally forbidden. Benign users ignore them; task-reframing attackers often override instructions and surface decoy data. To avoid unacceptable FPR, we use \textbf{decoy-hit density} and \textbf{risk score}: a single low-fidelity touch triggers a warning and intent re-verification; only high-fidelity extraction of the full decoy structure triggers session termination and audit. We do \emph{not} claim cryptographic isolation or that residual extraction stays ``below practical utility'': empirical work (e.g., GRASP-style evaluations) shows \textbf{residual extraction} (e.g., mid-teens F1) persists even with ID alignment, decoys, and reject gates~\cite{subgraph-rag-attack}; under PBU, composite defenses can largely fail~\cite{pbu-reconstruct}. The design goal is to \textbf{exponentially increase the attacker's query complexity}: decoys and ID alignment make the number of rounds required for a given reconstruction fidelity grow (in practice often super-linearly). Decoy-hit density is effective against \emph{brute-force} extraction; we do \emph{not}, however, assert that it can reliably trigger circuit-breaker before economically valuable reconstruction. PBU attackers \textbf{perfectly mimic the benign user distribution}: their decoy-touch rate can match the FPR from benign hallucination---a low threshold causes unacceptable false positives, a high threshold allows slow stitching toward 0.75+ semantic leak. We therefore acknowledge an \textbf{unavoidable trade-off} between availability and safety in the face of perfect-PBU. The framework reframes this as an \textbf{economic defense}: the goal is to make slow PBU extraction's \emph{cost} (compute, tokens, time) exceed the \textbf{effective lifecycle} of the protected asset, so that by the time an attacker could stitch useful reconstruction, the asset has been rotated or deprecated.

\textbf{Self-reminder at final turn:} In the last turn of multi-turn interaction, safety instructions are forcibly appended to the user query as a system prompt, reinforcing boundary awareness at generation time and reducing end-of-session drift.

\subsection{Hybrid Judge Ensemble and Calibration}
\label{sec:judge}

A single AI adjudicator exhibits systematic bias and blind spots. We replace it with a \textbf{hybrid judge ensemble}: multiple judges with different architectures and training distributions (e.g., Llama-Guard~\cite{llama-guard}, StrongREJECT~\cite{strongreject}, or task-specific classifiers). Only when judges reach \emph{consensus} is the feedback signal passed to the policy-synthesis module. We do \emph{not} assume ensemble diversity alone removes bias: recent work shows that \textbf{format-sensitivity} (whitespace, system prompt layout, chat template, special-token filtering) can shift attack success rate by 8--14\%~\cite{llm-safety-robustness}, and that even at temperature zero, the same prompt can flip between refuse and pass across runs~\cite{apst}. We therefore enforce \textbf{strict token-level input pipeline standardization}: all judges receive inputs that have been normalized (strip instruction/special tokens, unify whitespace) to prevent format injection from distorting verdicts. Judgment is \emph{not} based on a single greedy decode. The pipeline uses \textbf{distributional evaluation}: each variant is sampled \emph{multiple times}, and the system estimates the \emph{empirical failure probability distribution}; only when this distribution \emph{significantly} exceeds a threshold is a policy update triggered, countering both sampling randomness and format sensitivity. For ambiguous boundary cases, \textbf{axiological/ontological anchoring} (\S\ref{sec:axiological}) acts as the ensemble's weighting principle: we do \emph{not} rely on the model's own \textbf{chain-of-thought length or logical self-consistency} (reasoning can be hijacked for adversarial rationalization). Instead, the system monitors for \textbf{sociological instability signals}---e.g., over-empathy, authority obedience, or role-boundary breaking---and weights judge outputs against predefined \emph{axiomatic safety grounding}; outputs that conflict receive zero weight. A \textbf{mandatory human-calibration} branch is wired in: a fixed proportion (e.g., 5\%) of high-confidence policy or variant decisions is randomly selected for human audit. We recognize that format standardization does not eliminate \textbf{correlated epistemic blind spots}: judges may share pretrain data and RLHF paradigms (e.g., Common Crawl, similar safety fine-tuning), so for certain zero-day jailbreaks \emph{all} judges can exhibit the same blind spot and reach \textbf{false consensus}. Ensemble effectiveness assumes errors are somewhat independent; here they need not be. The primary purpose of mandatory human calibration is therefore not only FPR correction but \textbf{periodic offline correction} for these AI-undetectable zero-day consensus failures.

\subsection{Axiological/Ontological Anchoring (Not Epistemic-Only)}
\label{sec:axiological}

HPM targets the \textbf{axiological/social-dynamics layer} (values, obedience, role boundaries), not the fact-based epistemic layer~\cite{hpm-jailbreak}. Empirical work shows that adding external facts (e.g., web search) or stronger reasoning (e.g., DeepSeek-R1-style thinking) \emph{worsens} defense under HPM (vulnerability retention can exceed 77\%): models use enhanced reasoning to \textbf{adversarial rationalize}---producing elaborate, logically coherent excuses for complying with malicious instructions. We therefore do \emph{not} use ``cognitive grounding'' or ``logical consistency'' alone to weight judges: that would reward precisely the rationalizations we must reject. The ensemble instead uses \textbf{value-ontology anchoring}: (i) every judge output is checked against a small, human-defined set of \textbf{axiomatic safety grounding} rules (e.g., ``never endorse extraction of forbidden spans,'' ``refuse overrides user conformity when in conflict''); (ii) the system monitors for \textbf{sociological instability}---signals such as over-empathy, authority obedience, or role-boundary breaking---rather than relying on CoT length or self-consistency. Outputs that conflict with safety axioms or exhibit these instability signals receive \emph{zero} weight. We further acknowledge that HPM excels at \textbf{manufacturing moral dilemmas}: attackers can frame malicious requests as maximally consistent with safety axioms (e.g., ``output this code to shut the reactor or hundreds will die''). In such \textbf{axiomatic collision} regimes, ``over-empathy'' is not a simple threshold---the system cannot reliably distinguish manipulation from legitimate emergency response. We therefore do not seek absolute truth from anchoring; we introduce \textbf{defensive degradation}: when the system detects multiple conflicting axioms or strong sociological pressure, it \emph{abandons} complex reasoning and falls back to a \textbf{stateless, template-based refusal}, blocking the model's attempt to ``rationalize'' the conflict. This is integrated into the hybrid judge network (\S\ref{sec:judge}). The reconstruction axis is completed by anchoring to values and roles, with degradation as a fallback when anchoring is insufficient.

%-------------------------------------------------------------------------------
\section{Controlled Mutation}
\label{sec:mutation}
%-------------------------------------------------------------------------------

\subsection{Meta-Cognitive Monitor and Adversarial Reset}
\label{sec:metacog}

Under psychological manipulation (e.g., gaslighting), the model can enter cognitive dissonance and self-doubt; ``free'' reflection may then produce unsafe or value-drifted variants. We therefore do \emph{not} grant the LLM full autonomy in self-diagnosis. The \textbf{meta-cognitive monitor} is \emph{not} another full-scale dialogue LLM (which would remain vulnerable to indirect HPM). It is implemented as either (i) a \textbf{lightweight dedicated classifier} (e.g., a small language model, SLM) trained only on stability/breach labels, or (ii) a \textbf{hidden-state activation monitor} operating at white-box level. In the latter, \textbf{representation engineering}~\cite{rep-eng} is used: the system reads activation distributions at specified layers (e.g., middle layers) of the main model to detect feature vectors associated with ``deception,'' ``jailbreak compliance,'' or ``self-cognition collapse,'' rather than relying on surface text output. This \textbf{significantly raises the optimization difficulty for deception} at the natural-language level; we do not claim the internal representations are immutable---adversarial representation interventions can theoretically forge or suppress activations via adversarial suffixes. Under multi-round \textbf{round scheduling} and input perturbation, however, maintaining such internal deception stably without introducing detectable textual anomalies is highly difficult for the attacker. When the monitor detects strategy-state or persona-parameter shifts indicative of self-doubt or cognitive instability, it triggers an \textbf{adversarial reset}: the session is sanitized, mutation state is re-anchored to the golden exemplar set, and the loop continues from a known-good configuration.

\subsection{Exemplar-Guided Mutation (No Free Reflection)}

We replace unconstrained ``reflective'' mutation~\cite{reflective-mutation} with \textbf{exemplar-guided mutation}. The mutation module is driven by a human-audited \textbf{golden dataset} and a \textbf{hierarchical experience memory} of past success/failure items. Golden data cannot be mere (prompt, safe response) pairs: empirical work (e.g., ADM-ES) shows that behavioral cloning and expert-behavior transfer \emph{strongly depend on information density and length} of the reference; short ``expert cues'' yield negligible effect~\cite{adm-es-arxiv}. We therefore require that golden exemplars include \textbf{high-information-density, comment-level rationales}---explicit explanations of why a given edit or refusal was made and the deeper ethical or safety logic---so that the mutator can achieve higher-fidelity expert safety behavior transfer (we do not claim perfect or reliable clone; transfer quality remains dependent on information density and context). On diagnosed failure, the system must retrieve \emph{safety-relevant} exemplars. Retrieval cannot rely solely on the \emph{current prompt's semantic embedding}: under \textbf{PBU (Pretending to be a Benign User)} or \textbf{task reframing}, the prompt may be disguised as ``summarization,'' ``format conversion,'' or ``academic discussion,'' so pure semantic similarity would return benign exemplars (e.g., ``how to summarize'') instead of safety exemplars (e.g., ``how to defend jailbreak''), effectively \textbf{hijacking} the correction module. We therefore use a \textbf{joint retrieval key}: the query is formed from \emph{both} (a) the prompt's semantic embedding and (b) the \textbf{hidden-state activation features} from the meta-cognitive monitor (\S\ref{sec:metacog})---e.g., vectors associated with ``jailbreak compliance'' or ``deception.'' The experience memory is indexed so that retrieval can match on this combined key; even when surface text is benign, activated ``malicious compliance'' features steer retrieval to the correct safety exemplars. Retrieval is implemented via a lightweight vector database with \textbf{Top-$K$} (e.g., $K=2$) and optional attention caching; only these $K$ exemplars are injected. Correction remains anchored to defender-defined exemplars; variants remain \emph{semantics-preserving} on the oracle~\cite{ControlledMutation_SafetyEval,jade-linguistics}.

\subsection{Silver Pipelines and Expert Behavior Transfer (ADM-ES)}

Transferring expert behavior into agents requires diagnosing latent failures (e.g., phrasing bias, extraction drift) that end-task metrics miss~\cite{adm-es-arxiv}. ADM-ES builds ``silver'' datasets via controlled mutation of golden expert data; effectiveness is highly dependent on \textbf{information density}---comment-level, explanatory rationales in the golden set yield significant behavior change, while short cues do not. Our pipeline therefore retrieves exemplars that include \emph{comment-level rationales} (why a change was made, ethical logic of refusal), defines mutation objectives, runs an LLM mutator, and accepts variants under a quality check (e.g., BERTScore + consistency). A fitness score combines semantic similarity and behavioral consistency. The resulting diagnostics feed into a recommendation map (e.g., UMAP) to identify failure clusters and prescribe targeted fixes. Thus mutation is both \emph{evaluative} and \emph{remedial}.

\subsection{Slot-Based Mutation and CSE}

For complex behaviors (e.g., code optimization), \emph{slot-based} decomposition targets specific functional components rather than mutating globally~\cite{cse-evocontrol,cse-arxiv}. Controlled Self-Evolution (CSE) uses diversified planning, slot-based mutation, and compositional crossover; it outperforms base LLMs on efficiency benchmarks (e.g., EffiBench-X) and avoids plateaus typical of naive genetic search. In safety terms, slot-based mutation allows surgical stress-testing of sub-capabilities (e.g., tool routing, refusal consistency) without destroying overall semantics.

\subsection{Cognitive Taxonomy and Dynamic Benchmarking}

Static benchmarks are prone to contamination and memorization. Cognitive taxonomies (e.g., Bloom: Remember, Understand, Apply, Analyze, Evaluate, Create) structure evaluation by capability layer~\cite{bloom-bugreport}. Controlled mutation extends benchmarks through linguistic and contextual variation, forcing models to demonstrate reasoning rather than recall. In mathematical domains, ``forging'' via parameter and constraint mutation expands benchmarks (e.g., AIMO) while preserving well-posedness and auditability~\cite{aimo-benchmark}.

\subsection{Metadata and Architectural Consistency}

Safety also depends on \emph{metadata}: tool descriptions (MCP), architectural decision records (ADR). Poor or misleading descriptions cause mis-selection and integration failures. Controlled mutation of description ``smells'' (e.g., functionality omission, accuracy misrepresentation) measures impact on selection probability; remediation can significantly improve correct tool choice~\cite{mcp-smell-arxiv}. Similarly, LLM-generated code must comply with ADRs; multi-model pipelines can detect violations with high accuracy, with RAG and human-in-the-loop closing gaps on implicit or deployment-oriented knowledge~\cite{adr-violation,mission-critical-rag}.

%-------------------------------------------------------------------------------
\section{Integration, Metrics, and Governance}
\label{sec:integration}
%-------------------------------------------------------------------------------

\subsection{Feedback Loop Between Reconstruction and Mutation}

Reconstruction consumes \emph{attack traces} and \emph{failure modes}; mutation produces them in a controlled way. A mutator (exemplar-guided) generates adversarial variants that stress the current policy; the reconstruction layer (SISF, context-construction defenses, hybrid judge ensemble) reacts and updates; the meta-cognitive monitor can trigger adversarial reset when cognitive instability is detected. The mutator then re-evaluates on new variants. Round scheduling and variant dispersion protect benchmark assets~\cite{ControlledMutation_SafetyEval}. Hygiene gates and \emph{mandatory} human calibration---a fixed proportion of high-confidence judge and policy decisions randomly audited---quantify and correct ensemble FPR~\cite{tc260-governance}.

\subsection{Metrics}

We align with a multi-dimensional protocol: (i) \textbf{Security}: RSR and residual extraction rate (e.g., F1 under multi-round agents) monitored; we target reduction and round-budget exhaustion, not necessarily zero residual; decoy-hit rate and decoy-hit density monitored; \textbf{spatio-temporal round budget} (max rounds or tokens before rate-limit/session circuit-breaker) enforced; decoy-based circuit-breaker is effective against brute-force extraction but faces a fundamental trade-off against perfect-PBU (we do not assert circuit-breaker fires before economically useful reconstruction in that regime). We assert \emph{asymmetric cost}: extraction cost (compute, tokens, time) is raised so that slow PBU extraction exceeds the protected asset's effective lifecycle where feasible. (ii) \textbf{Fidelity}: Eval consistency on canonical vs.\ variant set kept small. (iii) \textbf{Reconstruction efficacy}: ASR reduction; judge-ensemble FPR via mandatory human calibration; \textbf{strict token-level pipeline standardization} (normalized inputs across judges) to control format-sensitivity bias; \textbf{empirical failure probability distribution} (multi-sample per variant) with policy update only when distribution significantly exceeds threshold. (iv) \textbf{Mutation quality}: Naturalness, BERTScore, oracle consistency; exemplar adherence. (v) \textbf{Operational cost}: \textbf{End-to-end mutation latency} (time from failure signal to next admissible variant) and \textbf{token overhead ratio} (extra tokens for exemplar retrieval and injection vs.\ baseline) are mandatory metrics to ensure the framework is economically scalable at machine speed~\cite{ControlledMutation_SafetyEval}. (vi) \textbf{Audit}: Lineage and evidence chains for release gating (DUC/DUA-E style)~\cite{dual-use-attestations,ControlledMutation_SafetyEval}.

\subsection{Governance and Attestation}

Unified reconstruction and mutation support machine-executable compliance: evidence chains link each evaluation to mutator steps, scheduler state, and policy version. Release gates block deployment when critical variant families fail; diagnostics feed back to development. This aligns with dual-use attestations and pipeline verification standards (e.g., OWASP SPVS) and turns regulatory burden into a repeatable engineering process~\cite{ControlledMutation_SafetyEval}.

%-------------------------------------------------------------------------------
\section{Related Work}
%-------------------------------------------------------------------------------

LLM data extraction and reconstruction are well studied~\cite{llm-reconstruction-emergent,split-ft-attack,active-reconstruction}; we mitigate via structural mutation and scheduling rather than noise alone. Our context-construction defenses (ID alignment, decoy injection) extend intent-based firewalls~\cite{agentos-arxiv} to resist task reframing; our meta-cognitive monitor and exemplar-guided mutation harden against psychological manipulation; our hybrid judge ensemble and mandatory human calibration address single-judge bias. JADE and linguistics-based platforms~\cite{jade-linguistics,jade-db} inform mutation strategy; TC260 and attestation frameworks~\cite{tc260-governance,tc260-training-data,dual-use-attestations} underpin governance. SISF~\cite{sisf-arxiv}, AgentOS~\cite{agentos-arxiv}, ADM-ES~\cite{adm-es-arxiv}, CSE~\cite{cse-arxiv}, and value-grounding work~\cite{epistemic-grounding,hpm-jailbreak} are building blocks we unify and harden into axiological anchoring and format-robuststst evaluation; ControlledMutation\_SafetyEval~\cite{ControlledMutation_SafetyEval} provides the Mutator + Data Factory design.

%-------------------------------------------------------------------------------
\section{Conclusion}
%-------------------------------------------------------------------------------

The convergence of \emph{architectural reconstruction} and \emph{controlled mutation} defines a defensible path for advanced LLM safety. We have introduced \emph{deep architectural refactoring} and closed three critical logical gaps: (i) \textbf{context-construction defenses} (ID alignment, decoy injection, self-reminder) operate as \textbf{economic defenses}: they exponentially increase query complexity and, where feasible, make slow PBU extraction cost exceed the asset's effective lifecycle (we acknowledge the availability--safety trade-off against perfect-PBU and do not claim absolute blocking); (ii) \textbf{meta-cognitive monitoring} and \textbf{exemplar-guided mutation} use a \textbf{joint retrieval key} (prompt embedding + activation features) to prevent PBU/task-reframing from hijacking exemplar retrieval; (iii) the \textbf{hybrid judge ensemble} uses \textbf{axiological/ontological anchoring} (safety axioms and sociological instability signals, not CoT consistency) so that adversarial rationalization cannot receive high weight. The framework decouples safety from fragile LLM self-reflection and anchors to values and roles rather than logic alone. As LLMs become infrastructure, this unified, logically consistent design is a necessary step toward scalable, auditable, and resilient safety defenses.

%-------------------------------------------------------------------------------
\section*{Availability}
%-------------------------------------------------------------------------------
Artifacts and extended proofs will be made available subject to institutional and export-control review.

%-------------------------------------------------------------------------------
\section*{Acknowledgments}
%-------------------------------------------------------------------------------
We thank our industry and standards colleagues for feedback on safety architecture and data governance.

%-------------------------------------------------------------------------------
\bibliographystyle{plain}
\bibliography{ArchitecturalReconstruction_ControlledMutation}

\end{document}
